graphsim: An R package for simulating gene expression data from graph structures of biological pathways

### 2020-05-01 paper/paper.Rmd
correct figure 2-4 formatting for graph structures

plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node = 2, cex.arrow = 4, arrow_clip = 0.2)
mtext(text = "(b) Inhibiting pathway structure", side=1, line=2, at=0.075, adj=0.5, cex=1.75)
box()
```

@@

```{r simulation_activating, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a graph structure}. An example of a graph structure (a) that has been used to derive a relationship matrix (b), $\\Sigma$  matrix (c) and correlation structure (d) from the relative distances between the nodes. Non-negative values are coloured white to red from $0$ to $1$. This $\\Sigma$ matrix has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here genes with closer relationships in the pathway structure show a higher correlation between simulated values.'}
# activating graph
state <- rep(1, length(E(graph)))
plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node=2, cex.arrow=4, arrow_clip = 0.2,
              sub = "(a) Activating pathway structure", cex.sub = 1.75)
box()

@@

#plot adjacency matrix
heatmap.2(make_adjmatrix_graph(graph), scale = "none", trace = "none",
          col = colorpanel(3, "grey75", "white", "blue"),
          colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "Adjacency matrix", cex.sub = 1.75)

#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
          scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "(b) relationship matrix", cex.sub = 1.75)

#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = expression(paste("(c) ", Sigma, " matrix")), cex.sub = 1.75)

#plot simulated expression data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE, state = state)
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "(e) Simulated expression data (log scale)", cex.sub = 1.75)

#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
```

@@

#generate parameters for inhibitions
state <- c(1, 1, -1, 1, 1, 1, 1, -1)
plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node=2, cex.arrow=4, arrow_clip = 0.2,
              sub = "(a) Inhitibing pathway structure", cex.sub = 1.75)
box()

@@

#plot adjacency matrix
heatmap.2(make_adjmatrix_graph(graph), scale = "none", trace = "none",
          col = colorpanel(3, "grey75", "white", "blue"),
          colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "Adjacency matrix", cex.sub = 1.75)

# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = expression(paste("(c) ", Sigma, " matrix")), cex.sub = 1.75)

#plot simulated expression data
expr <- generate_expression(100, graph,  state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "(e) Simulated expression data (log scale)", cex.sub = 1.75)

#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
```

![\small \textbf{\textbf{Simulating expression from a biological pathway graph structure.}}
Simulating expression from graph structure with inhibitions.}} The graph structure (a) of a
known biological pathway, the TGF-$\beta$ receptor signaling activates SMADs (R-HSA-2173789),
was used to derive a relationship matrix (b), $\Sigma$ matrix (c) and correlation structure
(d) from the relative distances between the nodes. These values are coloured blue to red from
$-1$ to $1$. This has been used to generate a simulated expression dataset of 100 samples
(coloured blue to red from low to high) via sampling from the multivariate normal distribution.
Here modules of genes with correlated expression can be clearly discerned.](figure4.pdf){ width=95% #fig:simulation_smad label="fig:simulation_smad"}

@@

#plot adjacency matrix
heatmap.2(make_adjmatrix_graph(graph), scale = "none", trace = "none",
          col = colorpanel(3, "grey75", "white", "blue"),
          colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "Adjacency matrix", cex.sub = 1.75)

@@

# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = expression(paste("(c) ", Sigma, " matrix")), cex.sub = 1.75)

#plot simulated expression data
expr <- generate_expression(100, graph,  state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "(e) Simulated expression data (log scale)", cex.sub = 1.75)

#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
```

>>>>

plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node = 2, cex.arrow = 4, arrow_clip = 0.2)
mtext(text = "(b) Inhibiting pathway structure", side=1, line=3, at=0.075, adj=0.5, cex=1.75)
box()
```

@@

```{r simulation_activating, fig.align='center', fig.show='hold', fig.width='1.0\\linewidth', fig.height='1.0\\linewidth', out.height='.415\\linewidth', out.width='.415\\linewidth', fig.retina=10, fig.margin = FALSE, fig.ncol = 2, fig.cap = '\\textbf{Simulating expression from a graph structure}. An example of a graph structure (a) that has been used to derive a relationship matrix (b), $\\Sigma$  matrix (c) and correlation structure (d) from the relative distances between the nodes. Non-negative values are coloured white to red from $0$ to $1$. This $\\Sigma$ matrix has been used to generate a simulated expression dataset of 100 samples (coloured blue to red from low to high) via sampling from the multivariate normal distribution. Here genes with closer relationships in the pathway structure show a higher correlation between simulated values.'}
# activating graph
state <- rep(1, length(E(graph)))
plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Activating pathway structure", side=1, line=3, at=0.075, adj=0.5, cex=1.75)
box()

@@

#plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
          scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) relationship matrix", side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#simulated data
expr <- generate_expression(100, graph, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE, state = state)

#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)),
              sub = "(e) Simulated expression data (log scale)", cex.sub = 1.75)
```

@@

#generate parameters for inhibitions
state <- c(1, 1, -1, 1, 1, 1, 1, -1)
plot_directed(graph, state=state, layout = layout.kamada.kawai,
              cex.node=2, cex.arrow=4, arrow_clip = 0.2)
mtext(text = "(a) Inhibiting pathway structure", side=1, line=3, at=0.075, adj=0.5, cex=1.75)
box()

@@

# #plot relationship matrix
heatmap.2(make_distance_graph(graph, absolute = FALSE),
          scale = "none", trace = "none", col = colorpanel(50, "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(b) relationship matrix", side=1, line=3, at=0.075, adj=0.5, cex=1.75)

# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#simulated data
expr <- generate_expression(100, graph,  state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)

#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3, at=0.075, adj=0.5, cex=1.75)
```

@@

@@

# #plot sigma matrix
heatmap.2(make_sigma_mat_dist_graph(graph, state = state, 0.8, absolute = FALSE),
scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = expression(paste("(c) ", Sigma, " matrix")), side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#simulated data
expr <- generate_expression(100, graph,  state = state, cor = 0.8, mean = 0,
comm = FALSE, dist =TRUE, absolute = FALSE)

#plot simulated correlations
heatmap.2(cor(t(expr)), scale = "none", trace = "none", col = colorpanel(50, "blue", "white", "red"),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(d) Simulated correlation", side=1, line=3, at=0.075, adj=0.5, cex=1.75)

#plot simulated expression data
heatmap.2(expr, scale = "none", trace = "none", col = bluered(50),
colsep = 1:length(V(graph)), rowsep = 1:length(V(graph)))
mtext(text = "(e) Simulated expression data (log scale)", side=1, line=3, at=0.075, adj=0.5, cex=1.75)
```
